\chapter*{Abstract}
Learning about software modeling languages from documentation can be a difficult and confusing process, and many of the currently existing modeling language tutorials are only marginally better. At the same time, players of video games spend hours upon hours learning to play games that require mastering complex strategies and concepts, without losing motivation or interest. This success for motivating learning effort seen in many games, is in turn supporting an emerging trend of educational games, designed to teach a wide range of subject to people of all ages. This thesis presents an exploration of the principles and strategies used by video games to teach players their mechanics, and an attempt to use these principles to teach software modeling in an engaging way.

\noindent
Focusing on modeling with \gls{uml} activities in the context of Reactive Blocks, two different approaches for teaching the concepts of this topic are presented. The first approach is simply an ``improved'' tutorial, utilizing principles such as \emph{interactiveness} and \emph{context-sensitivity of information and instruction} to engage learners. The second approach is an educational game, adding \emph{immersion} and \emph{visualization} to the learning experience.

\noindent
The design and prototype implementation of both the interactive tutorial and the educational game, and the principles they are based on, are described in detail. Both prototypes are also evaluated with respect to these principles, focusing on their usability and teaching potential, with the support of data from user tests.
